---
title: "Excel Data Cleaning in Tips in R"
author: "David Keyes"
date: "2/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(lubridate)
```

# Intro

I'm working on creating R equivalents to the Excel data cleaning tips found [here](https://www.evalu-ate.org/wp-content/uploads/formidable/14/ExcelTips-2020.pdf).

# Identify all cells that contain a specific word or (short) phrase in a column with open-ended text

```{r}
text_data <- tribble(
 ~response,
 "I love excel for data",
 "Data cleaning is great",
 "This is fun"
)
```

Using square brackets around the letter D might remove the need to determine the case.

```{r}
text_data %>% 
  # mutate(response_lower_case = str_to_lower(response)) %>% 
  mutate(contains_data = case_when(
    str_detect(response, "[d]ata") ~ "Y",
    TRUE ~ "N"
  ))
```


# Identify and remove duplicate data

```{r}
names_data <- tribble(
  ~first, ~last,
  "Joe", "Smith",
  "Joe", "Smith",
  "Joe", "Jones"
)
```

```{r}
names_data %>% 
  distinct(first, last)
```

```{r}
names_data %>% 
  get_dupes(first, last)
```


# Identify the outliers within a data set (e.g., dates or grades)

```{r}
dates <- tribble(
  ~date,
  "1/10/2019",
  "3/21/2018",
  "4/3/2019",
  "11/16/2019",
  "5/6/2018"
)
```

```{r}
dates %>% 
  filter(str_detect(date, "2019"))
```

Try mdy. I think with lubridate you are showing it the format it is and it looked like a US format mdy.

```{r}
dates %>% 
  mutate(date = mdy(date)) %>% 
  mutate(observation_year = year(date)) %>% 
  filter(observation_year == 2019)
```


# Separate data from a single column into two or more columns (e.g., first and last names)

```{r}
names_data_2 <- tribble(
  ~name,
  "Dottie Smith",
  "Harold Vicker",
  "Gloria Perez"
)
```

```{r}
names_data_2 %>% 
  separate(name, into = c("first", "last"), sep = " ")
```


# Categorize data in a column (e.g., class assignments or subject groups)

```{r}
agreement_data <- tribble(
  ~response,
  "Agree",
  "Agree",
  "Strongly Disagree",
  "Strongly Agree",
  "Disagree",
  "Agree",
  "Agree",
  "Disagree",
  "Disagree",
  "Strongly Disagree",
  "Agree"
)
```

```{r}
agreement_data %>% 
  mutate(positive_negative = case_when(
    response %in% c("Agree", "Strongly Agree") ~ "Positive",
    response %in% c("Disagree", "Strongly Disagree") ~ "Negative"
  ))
```

