---
format:
  noaa-fisheries-memo-to-record-docx: default
execute: 
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk: 
    dev: ragg_png
    dpi: 300
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(flextable)
library(scales)
library(noaafisherieswcr)
library(janitor)
```

```{r}
options(scipen = 999)
```


**DATE:** [TODO]

**MEMORANDUM FOR:** [TODO]

**FROM:** [TODO]

**SUBJECT:** [TODO]

**ASSOCIATED DOCUMENTS OR DATA FILES:** [TODO]

## SUMMARY

[TODO]

### Variance-weighted Mean Survival

[TODO]

### Confidence Intervals for Juvenile Production Estimate

[TODO]

```{r}
jpe_data_preliminary <-
    calculate_jpe_data(
        JPI = 278448,
        V.JPI = 280340342,
        timing = "Preliminary"
    )

jpe_data_final <-
    calculate_jpe_data(
        JPI = 311058,
        V.JPI = 452638929,
        timing = "Final"
    )

jpe_data_preliminary |>
    bind_rows(jpe_data_final) |>
    pivot_wider(
        id_cols = Variable,
        names_from = timing,
        values_from = value
    ) |>
    filter(Variable %in% c(
        "f.hat",
        "natural surv rate",
        "JPI",
        "JPE",
        "JPE lower 95% CI",
        "JPE upper 95% CI"
    )) |>
    # slice(2, 1, 3, 4, 5) |> # Change if you want to reorder rows
    noaa_table() |>
    colformat_double(
        i = c(1, 4, 5, 6),
        j = c(2, 3),
        digits = 0,
        big.mark = ","
    )
```


## VWM survival and natural-origin JPE[^1]

[^1]: The Juvenile Production Estimate (JPE) reported here, and implemented in the package WCR.Fisheries.xxxx (available at <NMFS internal github>), uses the "Method 2" approach described in O'Farrell et al. 2018. Staff at the NOAA Fisheries Southwest Fisheries Science Center provide the latest estimates each year for outmigration survival (Arnold Amman) and the fry-to-smolt survival rate (Michael O'Farrell). O’Farrell M. R., W. H. Satterthwaite, A. N. Hendrix, and M. S. Mohr. 2018. Alternative Juvenile Production Estimate (JPE) forecast approaches for Sacramento River winter-run Chinook Salmon. San Francisco Estuary & Watershed Science 16(4):4.  https://doi.org/10.15447/sfews.2018v16iss4art4




### Preliminary estimate

```{r}
jpe_data_preliminary |>
    select(-timing) |>
    rename(`Preliminary Value` = value) |>
    noaa_table() |>
    colformat_double(
        i = c(1, 5, 6, 7, 10, 11),
        j = 2,
        digits = 0,
        big.mark = ","
    ) |>
    colformat_double(
        i = c(2, 4, 8, 9),
        j = 2,
        digits = 4,
        big.mark = ","
    )
```


### Final estimate

```{r}
jpe_data_final |>
    select(-timing) |>
    rename(`Final Value` = value) |>
    noaa_table() |>
    colformat_double(
        i = c(1, 5, 6, 7, 10, 11),
        j = 2,
        digits = 0,
        big.mark = ","
    ) |>
    colformat_double(
        i = c(2, 4, 8, 9),
        j = 2,
        digits = 4,
        big.mark = ","
    )
```


## Exploration of survival by year

```{r}
survival_by_year_tidy <-
    survival_by_year |>
    pivot_longer(cols = -year) |>
    separate_wider_delim(
        name,
        delim = ".",
        names = c(
            "measure",
            "location",
            "region"
        )
    ) |>
    filter(measure != "SE") |>
    mutate(location = str_to_title(location)) |>
    mutate(year = as.character(year))
```


```{r}
#| fig-height: 7
survival_by_year_tidy |>
    ggplot(aes(
        x = year,
        y = value,
        group = location
    )) +
    geom_col() +
    scale_y_continuous(
        labels = percent_format()
    ) +
    facet_wrap(
        vars(location),
        ncol = 1
    ) +
    labs(title = "Survival Rate by Location") +
    theme_minimal(
        base_family = "Arial",
        base_size = 12
    ) +
    theme(
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(size = 10),
        plot.title = element_text(
            size = 16,
            face = "bold"
        ),
        plot.title.position = "plot",
        panel.spacing = unit(50, "pt"),
        strip.text = element_text(
            hjust = 0,
            size = 12
        )
    )
```
