
Load packages:
```{r}
library(tidyverse)
library(geomtextpath)
```



Load vehicles data:
```{r}
#| message: false

vehicles <- fueleconomy::vehicles

avg_hwy <- vehicles |> 
  filter(year %in% c(1985:1990),
         drive %in% c("Rear-Wheel Drive", 
                      "4-Wheel or All-Wheel Drive")) |> 
  group_by(year, drive) |> 
  summarise(mean_hwy = mean(hwy))
```



Create a plot with a legend:
```{r}
avg_hwy |> 
  ggplot(aes(x = year, 
           y = mean_hwy, 
           color = drive)) +
  geom_line(linewidth = 0.75) +
  labs(title = "Average Highway Gas Mileage",
       x = NULL,
       y = NULL,
       color = "Year") +
  lims( y = c(18, 22)) +
  theme_minimal(base_size = 15)
```
Yikes! That's a huge legend!



Remove legend and add text line:
```{r}
avg_hwy |> 
  ggplot(aes(x = year, 
             y = mean_hwy, 
             color = drive,
             label = drive)) +              # added    
  geom_textline(linewidth = 0.75) +         # changed
  labs(title = "Average Highway Gas Mileage",
       x = NULL,
       y = NULL) +
  lims( y = c(18, 22)) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "none")           # added
```



Move text above line and increase font size:
```{r}
avg_hwy |> 
  ggplot(aes(x = year, 
             y = mean_hwy, 
             color = drive,
             label = drive)) +
  geom_textline(linewidth = 0.75, 
                vjust = -.25,               # added
                size = 5) +                 # added
  labs(title = "Average Highway Gas Mileage",
       x = NULL,
       y = NULL) +
  lims( y = c(18, 22)) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "none")
```



Smooth out text! 
Try some values between 0 & 100:
```{r}
avg_hwy |> 
  ggplot(aes(x = year, 
             y = mean_hwy, 
             color = drive,
             label = drive)) +
  geom_textline(linewidth = 0.75, 
                vjust = -0.25, 
                size = 5,
                text_smoothing = 40) +        # added
  labs(title = "Average Highway Gas Mileage",
       x = NULL,
       y = NULL) +
  lims( y = c(18, 22)) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "none")
```



What are the defaults for `geom_textline()`?
```{r}
GeomTextline$default_aes
```



So far, we've added annotations that look like `geom_text()` - what about `geom_label()`?
```{r}
avg_hwy |> 
  ggplot(aes(x = year, 
             y = mean_hwy, 
             color = drive,
             label = drive)) +
  geom_labelline(linewidth = 0.75,          # changed
                text_smoothing = 40, 
                size = 5) + 
  labs(title = "Average Highway Gas Mileage",
       x = NULL,
       y = NULL) +
  lims( y = c(18, 22)) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "none")
```

Next up, annotation lines!












Create fake data to play with annotation lines:
```{r}
break_even <- 
  tibble(
  units = seq(0, 100, 1),
  price = 50,
  fixed_cost = 750,
  variable_cost = 25,
  total_cost = fixed_cost + (variable_cost * units),
  total_revenue = price * units,
  break_even_units = fixed_cost / (price - variable_cost)
)
```



Plot two separate lines, revenue and cost:
```{r}
break_even |> 
  ggplot() +
  geom_line(aes(x = units, 
                y = total_cost),
            linewidth = 1,
            color = "red4",
            hjust = 0.7,
            size = 5) +
  geom_line(aes(x = units, 
                y = total_revenue),
            linewidth = 1,
            color = "green3",
            hjust = 0.7, 
            size = 5) +
  labs(title = "Break Even Analysis",
       x = "Units",
       y = "Total Cost and Revenue ($)") +
  theme_minimal(base_size = 15)
```
Notice there is no legend by default 🤔



Plot them using `geom_textline()` instead:
```{r}
break_even |> 
  ggplot() +
  geom_textline(aes(x = units, 
                y = total_cost,
                label = "Cost"),
            linewidth = 1,
            color = "red4",
            hjust = 0.7,
            size = 5) +
  geom_textline(aes(x = units, 
                y = total_revenue,
                label = "Revenue"),
            linewidth = 1,
            color = "green3",
            hjust = 0.7, 
            size = 5) +
  labs(title = "Break Even Analysis",
       x = "Units",
       y = "Total Cost and Revenue ($)") +
  theme_minimal(base_size = 15)
```



Add lines showing: 
- what our fixed costs are
- how many units we need to sell to break even
```{r}
break_even |> 
  ggplot() +
  geom_textline(aes(x = units, 
                y = total_cost,
                label = "Cost"),
            linewidth = 1,
            color = "red4",
            hjust = 0.7,
            size = 5) +
  geom_textline(aes(x = units, 
                y = total_revenue,
                label = "Revenue"),
            linewidth = 1,
            color = "green3",
            hjust = 0.7, 
            size = 5) +
  geom_texthline(aes(yintercept = fixed_cost, 
                     label = paste0("Fixed Cost = $", fixed_cost)),
                 color = "steelblue4",
                 linetype = "dotted",
                 linewidth = .75,
                 hjust = 0.6,
                 size = 4.5) +
  geom_textvline(aes(xintercept = break_even_units, 
                     label = paste0(break_even_units, " units")),
                 color = "steelblue4",
                 linetype = "dotted",
                 linewidth = .75,
                 hjust = 0.6,
                 size = 4.5) +
  labs(title = "Break Even Analysis",
       x = "Units",
       y = "Total Cost and Revenue ($)") +
  theme_minimal(base_size = 15)
```
Next up, just using the text without the line!












Load penguins data:
```{r}
penguins <- palmerpenguins::penguins |> 
  drop_na()
```



Density plot with `geom_textdensity()`:
```{r}
penguins |> 
  filter(species == "Adelie") |> 
  ggplot(aes(x = bill_length_mm, 
             color = island, 
             label = island)) +
  geom_textdensity(linewidth = 1, 
                   size = 4, 
                   hjust = .49,
                   spacing = 40) +
  labs(title = "Adelie Penguins from Different Islands",
       subtitle = "Do they have different bill lengths?",
       x = "Bill Length (mm)",
       y = "Density") +
  theme_minimal(base_size = 15) +
  theme(legend.position = "none",
        plot.subtitle = element_text(size = 14))
```
But I want a FILLED density plot!



Filled density plot with transparent colors:
```{r}
penguins |> 
  filter(species == "Adelie") |> 
  ggplot(aes(x = bill_length_mm, 
             fill = island,
             color = island,
             label = island)) +
  geom_density(alpha = .5) +
  scale_color_manual(values = c("#CEA804", "#EA5F94", "#0000FF")) +
  scale_fill_manual(values = c("#FFD700", "#EA5F94", "#0000FF")) +
  labs(title = "Adelie Penguins from Different Islands",
       subtitle = "Do they have different bill lengths?",
       x = "Bill Length (mm)",
       y = "Density",
       fill = "Island", 
       color = "Island") +
  theme_minimal(base_size = 15) +
  theme(plot.subtitle = element_text(size = 14))
```
That legend has to go.



Add floating annotations with `text_only = TRUE`:
```{r}
#| warning: false

penguins |> 
  filter(species == "Adelie") |> 
  ggplot(aes(x = bill_length_mm, 
             fill = island,
             color = island,
             label = island)) +
  geom_density(alpha = .5) +
  geom_textdensity(text_only = TRUE,
                   text_smoothing = 25,
                   hjust = "ymax",               # new!
                   vjust = -0.30, 
                   spacing = 75,                 # new!
                   size = 3.5) +
  scale_color_manual(values = c("#CEA804", "#EA5F94", "#0000FF")) +
  scale_fill_manual(values = c("#FFD700", "#EA5F94", "#0000FF")) +
  labs(title = "Adelie Penguins from Different Islands",
       subtitle = "Do they have different bill lengths?",
       x = "Bill Length (mm)",
       y = "Density") +
  ylim(c(0, .16)) +
  theme_minimal(base_size = 15) +
  theme(legend.position = "none",
        plot.subtitle = element_text(size = 14))
```
Happy plotting!



































