---
title: "Advanced tricks to put your data into the right format with `pivot_longer()` and `pivot_wider()`"
author: Albert Rapp
date: Nov 15, 2023
toc: false
format: gfm
wrap: none
knitr:
  opts_chunk:
    out.width: 70%
    collapse: true
    comment: "#>"
editor_options: 
  chunk_output_type: console
---

Last week, we started to learn about `pivot_longer()` and `pivot_wider()`.
These are two essential functions to speed up you data wrangling process.
Check out [the video and blog post from last week](https://rfortherestofus.com/2023/11/pivot-functions) if you haven't seen that yet.

As promised, this week we'll continue on this path and learn some of the advanced tricks that these two functions have to offer.
This should help you clean your data even faster.
Let's begin by revisiting what we did last week.

## Taylor Swift again

Last week, we worked with a data set from [tidyTuesday](https://github.com/rfordatascience/tidytuesday/blob/master/data/2023/2023-10-17/readme.md).
And not just any data set, we worked with this data set about Taylor Swift albums.

```{r}
#| message: false
#| warning: false
library(tidyverse)
taylor_albums <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-10-17/taylor_albums.csv') |> 
  filter(!ep)
taylor_albums 
```

We figured out that we can rearrange the data from the columns `metacritic_score` and `user_score` so that we can pass this new data set to `ggplot()` for quick plotting. 

```{r}
taylor_longer <- taylor_albums |> 
  pivot_longer(
    cols = c(metacritic_score, user_score),
    names_to = 'score_type',
    values_to = 'score'
  )
taylor_longer

taylor_longer |> 
  ggplot(aes(x = album_release, y = score, color = score_type)) +
  geom_line() +
  geom_point() +
  facet_wrap(vars(score_type), scales = 'free_y') +
  theme_minimal(base_size = 14, base_family = 'Source Sans Pro') +
  labs(
    x = element_blank(),
    y = 'Score',
    color = 'Type'
  )
```

## Nicer labels

Notice that in our previous chart, the labels are a bit redundant.
We always write `metacritic_score` or `user_score`.
Why not just `Metacritic` or `User` (spelled with a capital letter)?

Well, we could manually make this look nicer.
But this would require working with the text variables using functions like `str_remove_all()` or `str_to_title()`.

```{r}
taylor_longer |> 
  mutate(
    score_type = score_type |> str_remove_all('_score') |> str_to_title() 
  )
```

See how the labels in the `score_type` column now say what we'd want to show in a ggplot?
That's great.
We could pass this to `ggplot()` now and everything would be fine.
But all of this was an extra step we had to do.
Can't we just let `pivot_longer()` handle that as it's rearranging the data?

Well, we're in luck.
It turns out `pivot_longer()` can do all of this for us.
The trick here is to also specify the arguments `names_pattern` and `names_transform`.
Here's what they do.

- `names_pattern`: Describes a so-called *regular expression (regex)* that describes the pattern of the column names and by specifying groups with `()` we can tell `pivot_longer()` which parts we want to extract. 

- `names_transform`: Describes a function that transforms the labels in the end. In our case this could just be `str_to_title` (without paranthesis).

```{r}
taylor_albums |> 
  pivot_longer(
    cols = c(metacritic_score, user_score),
    names_to = 'score_type',
    values_to = 'score',
    names_pattern = '(.+)_score',
    names_transform = str_to_title
  )
```

Neat, this worked out pretty nicely.
But what´s that `(.+)` we used?
Here, this is part of the regular expression we built. 
If you're not familiar, regular expressions are a way to describe complex text patterns. 
They use an unusual syntax (described below), but are very powerful once you learn to use them (they are covered extensively in our [Data Cleaning with R course](https://rfortherestofus.com/courses/data-cleaning/)).
So, without going into too much details about regex in general, let's go through what we did here one by one.

- `.`: This is a placeholder that can mean any character (except for a new line)
- `+`: This means that whatever preceded this symbol, it can show up once or multiple times (but at least once).
- `.+`: Together this means that this will "catch" any text that consist out of anything but a line break
- `.+_score`: This means that this catches all patterns that consists out of text without line breaks that are followed by the text `_score`. This means that our regex describes the exact pattern that our column names `metacritic_score` and `user_score` have.
- `(.+)_score`: Adding the parentheses tells `pivot_longer()` which part of the pattern we are interested in. Here that's what comes before `_score`.


Oof. 
That was a lot to digest, I know.
You may wonder why it's worth figuring this stuff out.
This technique really shines with more complex data sets that you may find in the wild.
Let's have a look.


## A more complex example

Here's another data set from TidyTuesday.
It's about the wages of nurses in different states of the US.

```{r}
nurses <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-10-05/nurses.csv') |>  
  janitor::clean_names()
nurses
```


As you can see, this data set has a loooot of columns.
It's pretty wide, you might say.
Let's check out how wide by just considering the column names.

```{r}
colnames(nurses)
```

Let's narrow this down a little bit.
We have a lot of columns about average, median and [percentiles](https://en.wikipedia.org/wiki/Percentile) of hourly and annual salary.
We can select only those columns by using the tidyselect helper `matches()`.
In there, we have to specify that we are looking for columns with the word 'hourly' or 'annual' in them.
That's done with the `|` operator (another regex by the way).

```{r}
nurses |> 
  select(state, year, matches('hourly|annual')) |> 
  colnames()
```

Now look at those names.
Do you see a pattern?
Apart from the column `state` and `year`, we always have the following pattern

- hourly or annual,
- an underscore `_` and
- one of the following words:
    - "wage_avg", 
    - "wage_median", 
    - "salary_avg", 
    - "salary_median", 
    - some number followed by "th_percentile"


This means that from each column we can actually extract two information:

- Are we talking about hourly or annual salary?
- What kind of quantity of that salary do we mean: average, median or some other percentile?

Luckily, we can catch all of this with one regex that contains **two** groups (indicated by `()`).
Here's how that could look in `pivot_longer()`.

```{r}
nurses |> 
  select(state, year, matches('hourly|annual')) |> 
  pivot_longer(
    cols = -c(state, year),
    names_pattern = '(.+)_(.+)',
    names_to = c('timeframe', 'type'),
    values_to = 'wage'
  )
```

Oh no.
It seems like the `timeframe` column contains more than just "hourly" or "annual".
That's because our regex `(.+)_(.+)` was a bit ambiguous.
Since the `.` operator catches all things including underscores `_` it is not clear weather `pivot_longer()` should split at the first or second underscore.

But we can fix that.
Instead of using `.` in the first group, we can use `[a-z]`.
This means what we only want to "catch" things that contain lower letters a-z.
As the percentiles also contain numbers (as in `25th_percentile`) we get an unambiguous split.

```{r}
nurses_longer <- nurses |> 
  select(state, year, matches('hourly|annual')) |> 
  pivot_longer(
    cols = -c(state, year),
    names_pattern = '([a-z]+)_(.+)',
    names_to = c('timeframe', 'type'),
    values_to = 'wage'
  )
nurses_longer
```

And with that data set we could now take a look at a specific state in a chart.

```{r}
#| fig-width: 12
#| fig-height: 10
#| out-width: 100%
nurses_longer |> 
  filter(state == 'Alabama') |> 
  ggplot(aes(x = year, y = wage, group = type)) +
  geom_line() +
  geom_text(
    data = nurses_longer |> filter(state == 'Alabama', year == 2020),
    aes(label = type),
    hjust = 0,
    nudge_x = 0.5,
    size = 6,
    family = 'Source Sans Pro'
  ) +
  facet_wrap(vars(timeframe), scales = 'free_y') +
  theme_minimal(base_size = 24, base_family = 'Source Sans Pro') +
  coord_cartesian(xlim = c(1998, 2030)) +
  labs(
    x = element_blank(),
    y = 'Salary',
    title = 'Salary of Nurses in Alabama'
  )
```

# Conclusion

Of course, there's lots more to do to polish our nurses chart.
But the first step here was (once again) getting the data into the right format.
In this blog post, we have seen that we can use pretty advanced tricks like regex to let `pivot_longer()` (and similarly `pivot_wider()`) rearrange the data.

Clearly, these advanced steps require a bit getting used to.
So don't worry if you don't get it immediately.
And with that said, I'll give you a little bit of time to think these things through and then I'll see you next week 👋





