Load packages:
```{r}
library(fs)
library(stringr)
```

Create a directory for this demo:
```{r}
dir_create("fs-file-management/fs-example")
dir_exists("fs-file-management/fs-example")
```

Create some poorly named files:
```{r}
files <- c(
  "RAW-DATA.csv",
  "Clean Data.csv",
  "1-analysis.R",
  "2_visualization.R",
  "Very important report.qmd",
  "Very important report.pdf",
  "Very important report_final.pdf"
)

file_create("fs-file-management/fs-example", files)
```

Built in safety to not overwrite directories and files: 
```{r}
#| error: true
file_create("fs-file-management/fs-example", "1-analysis.R")

dir_create("fs-file-management/fs-example")
```

Demonstrate the directory tree:
```{r}
dir_tree("fs-file-management/fs-example")
```

Get file names:
```{r}
files_old <- dir_ls("fs-file-management/fs-example")

files_old
```

Standardize file names in a new vector:
```{r}
files_new <- files_old |>
  str_to_lower() |>
  str_replace_all(" ", "-") |>
  str_replace_all("_", "-")

files_new
```

Rename the old file names with the new names:
```{r}
file_move(files_old, files_new)

dir_ls("fs-file-management/fs-example")
```

Create subdirectories:
```{r}
subdirs <- c("R", "data", "reports")

dir_create("fs-file-management/fs-example", subdirs)

dir_tree("fs-file-management/fs-example")
```

Get vectors of the same file type:
```{r}
r_files <- dir_ls("fs-file-management/fs-example", glob = "*.r|*.R")
data_files <- dir_ls("fs-file-management/fs-example", glob = "*.csv")
reports <- dir_ls("fs-file-management/fs-example", glob = "*.qmd|*.pdf")
```

Move files into the new subdirectories:
```{r}
file_move(r_files, "fs-file-management/fs-example/R")
file_move(data_files, "fs-file-management/fs-example/data")
file_move(reports, "fs-file-management/fs-example/reports")
```

Check it worked as expected:
```{r}
dir_tree("fs-file-management/fs-example")
```

Clean up and delete our demo directory:
```{r}
dir_exists("fs-file-management/fs-example")

dir_delete("fs-file-management/fs-example")

dir_exists("fs-file-management/fs-example")
```

Automate the process with custom functions:

```{r}
dir_create("fs-file-management/fs-example")

files <- c(
  "RAW-DATA.csv",
  "Clean Data.csv",
  "1-analysis.R",
  "2_visualization.R",
  "Very important report.qmd",
  "Very important report.pdf",
  "Very important report_final.pdf"
)

file_create("fs-file-management/fs-example", files)
```

```{r}
#| eval: false
clean_file_names <- function(folder) {
  files_old <- dir_ls(folder)

  files_new <- files_old |>
    str_to_lower() |>
    str_replace_all(" ", "-") |>
    str_replace_all("_", "-")

  file_move(files_old, files_new)
}

# Run the function
clean_file_names("fs-file-management/fs-example")
```

```{r}
#| eval: false
organize_files <- function(folder) {
  subdirs <- c("R", "data", "reports")

  dir_create(folder, subdirs)

  r_files <- dir_ls(folder, glob = "*.r|*.R")
  data_files <- dir_ls(folder, glob = "*.csv|*.xlsx")
  reports <- dir_ls(folder, glob = "*.qmd|*.pdf|*.rmd")

  file_move(r_files, path(str_glue("{folder}/R")))
  file_move(data_files, path(str_glue("{folder}/data")))
  file_move(reports, path(str_glue("{folder}/reports")))
}

# Run the function
organize_files("fs-file-management/fs-example")

# See the results
dir_tree("fs-file-management/fs-example")
```

Clean up and delete our demo directory:
```{r}
dir_delete("fs-file-management/fs-example")
```