# Load Packages

Let's load the two packages we'll use

```{r}
library(tidyverse)
library(scales)
```

# Generate Data

We'll create some fake data. 

```{r}
set.seed(1234)

school_quality <-
  tibble(
    id = seq(1, 300, 1),
    school = rep(c(
      "Sabin", "Vernon", "Faubion", "Irvington", "Alameda", "Beverly Cleary"
    ), 50),
    opinion = sample(c("Very bad", "Bad", "Good", "Very Good"), 300, replace = TRUE)
  )

school_quality
```

# Default Bar Chart

Let's make a summary of our data. 

```{r}
school_quality_summary <- school_quality %>% 
  group_by(school, opinion) %>% 
  count(name = "n_answers") %>% 
  group_by(school) %>% 
  mutate(percent_answers = n_answers / sum(n_answers)) %>% 
  ungroup() %>% 
  mutate(percent_answers_label = percent(percent_answers, accuracy = 1))

school_quality_summary
```


```{r}
school_quality_summary %>%
  ggplot(aes(x = school, 
             y = percent_answers,
             fill = opinion)) +
  geom_col() +
  geom_text(
    aes(label = percent_answers_label),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold"
  ) +
  coord_flip() +
  scale_x_discrete() +
  scale_fill_viridis_d() +
  labs(title = "How good is the education at your school?",
       x = NULL,
       fill = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        panel.grid = element_blank(),
        legend.position = "top")
```


# Basic Diverging Bar Chart

Let's create data that we need to make a basic diverging bar chart.

```{r}
school_quality_summary_diverging <- school_quality_summary %>%
  mutate(percent_answers = if_else(opinion %in% c("Very Good", "Good"), percent_answers, -percent_answers)) %>% 
  mutate(percent_answers_label = percent(percent_answers, accuracy = 1))

school_quality_summary_diverging
```
We can now make a basic diverging bar chart. 

```{r}
school_quality_summary_diverging %>%
  ggplot(aes(x = school, 
             y = percent_answers,
             fill = opinion)) +
  geom_col() +
  geom_text(
    aes(label = percent_answers_label),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold"
  ) +
  coord_flip() +
  scale_x_discrete() +
  scale_fill_viridis_d() +
  labs(title = "How good is the education at your school?",
       x = NULL,
       fill = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        panel.grid = element_blank(),
        legend.position = "top")
```


# Positive/Negative Labels

In our previous chart, the labels for very bad and bad were negative. Let's change this!

This will make our labels are positive numbers. 

```{r}
school_quality_summary_diverging_good_labels <- school_quality_summary_diverging %>%
  mutate(percent_answers_label = abs(percent_answers)) %>% 
  mutate(percent_answers_label = percent(percent_answers_label, accuracy = 1))

school_quality_summary_diverging_good_labels
```

```{r}
school_quality_summary_diverging_good_labels %>% 
  ggplot(aes(x = school, 
             y = percent_answers,
             fill = opinion)) +
  geom_col() +
  geom_text(
    aes(label = percent_answers_label),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold"
  ) +
  coord_flip() +
  scale_x_discrete() +
  scale_fill_viridis_d() +
  labs(title = "How good is the education at your school?",
       x = NULL,
       fill = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        panel.grid = element_blank(),
        legend.position = "top")
```


# Reorder Bars

Our bars are out of order. Let's fix this!

```{r}
school_quality_summary_diverging_right_order <- school_quality_summary_diverging_good_labels %>% 
  mutate(opinion = fct_relevel(opinion,
                               "Bad", "Very bad", "Good", "Very Good"),
         opinion = fct_rev(opinion)) 

school_quality_summary_diverging_right_order
```


```{r}
school_quality_summary_diverging_right_order %>%
  ggplot(aes(x = school, 
             y = percent_answers,
             fill = opinion)) +
  geom_col() +
  geom_text(
    aes(label = scales::percent(abs(percent_answers), accuracy = 1)),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold"
  ) +
  coord_flip() +
  scale_x_discrete() +
  scale_fill_viridis_d() +
  labs(title = "How good is the education at your school?",
       x = NULL,
       fill = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        panel.grid = element_blank(),
        legend.position = "top")
```

# Make Legend Order Match

The bars are now in the right order, but the legend doesn't match. Let's fix this!

```{r}
school_quality_summary_diverging_right_order %>%
  ggplot(aes(x = school, 
             y = percent_answers,
             fill = opinion)) +
  geom_col() +
  geom_text(
    aes(label = percent_answers_label),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold"
  ) +
  coord_flip() +
  scale_x_discrete() +
  scale_fill_viridis_d(breaks = c("Very bad", "Bad", "Good", "Very Good")) +
  labs(title = "How good is the education at your school?",
       x = NULL,
       fill = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        panel.grid = element_blank(),
        legend.position = "top")
```



# Improve Colors

Let's use a more appropriate color scale for this data. 

```{r}
school_quality_summary_diverging_right_order %>%
  ggplot(aes(x = school, 
             y = percent_answers,
             fill = opinion)) +
  geom_col() +
  geom_text(
    aes(label = percent_answers_label),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold"
  ) +
  coord_flip() +
  scale_x_discrete() +
  scale_fill_manual(breaks = c("Very bad", "Bad", "Good", "Very Good"),
                    values = c(
                      "Very bad" = "darkorange3",
                      "Bad" = "orange",
                      "Good" = "deepskyblue",
                      "Very Good" = "deepskyblue4"
                    )) +
  labs(title = "How good is the education at your school?",
       x = NULL,
       fill = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        panel.grid = element_blank(),
        legend.position = "top")
```


